---
title: "Make Countries Move"
author: "Adela Sobotkova"
date: "05/1/2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#  Explore global development with R
In this exercise, you will load a filtered `gapminder` dataset -  with a subset of data on global development from 1952 - 2007 in increments of 5 years - to capture the period between the Second World War and the Global Financial Crisis. 

**Your task: Explore the data and visualise it in both static and animated ways, providing answers and solutions to 7 questions/tasks within this script.** 

- **Submit to Github this script with solutions and its knitted version.**  
- **Submit to Brightspace a document with the (1) URL of this script and its knitted version in Github (2) answer of 250 words and a (3) visualisation of the Final Question**

## Get the necessary packages
First, start with installing and activating the relevant packages `tidyverse`, `gganimate`, and `gapminder` if you do not have them already. 
*Pay attention to what warning messages you get when installing `gganimate`, as your computer might need other packages than `gifski` and `av`*

```{r libraries, echo = FALSE}
# install.packages("gganimate")
# install.packages("gifski")
# install.packages("av")
# install.packages("gapminder")
library(tidyverse)
library(gganimate)
library(gifski)
library(av)
library(gapminder)

```

## Look at the data and tackle the tasks
First, see which specific years are actually represented in the dataset and what variables are being recorded for each country. Note that when you run the cell below, Rmarkdown will give you two results - one for each line - that you can flip between.
```{r}
str(gapminder)
unique(gapminder$year)
?gapminder
data(gapminder)
head(gapminder)
```

The dataset contains information on each country in the sampled year, its continent, life expectancy, population, and GDP per capita.

Let's plot all the countries in 1952.
```{r 1952}
theme_set(theme_bw())  # set theme to white background for better visibility

options(scipen = 999)

ggplot(subset(gapminder, year == 1952), aes(gdpPercap, lifeExp, size = pop)) +
  geom_point(aes(color = continent)) +
  scale_x_log10(labels = scales::comma) +
  scale_size(labels = scales::comma)+
  ggtitle("World in 1952") +
  labs(x="GDP per capita",
       y= "life expectancy") +
  theme(text = element_text(size = 14))
```

...

We see an interesting spread with an outlier to the right. Explore who it is so you can answer question 2 below!


Next, you can generate a similar plot for 2007 and compare the differences
```{r 2007}
ggplot(subset(gapminder, year == 2007), aes(gdpPercap, lifeExp, size = pop)) +
  geom_point() +
  scale_x_log10() +
  ggtitle("Figure 02")

options(scipen = 999)

ggplot(subset(gapminder, year == 2007), aes(gdpPercap, lifeExp, size = pop)) +
  geom_point(aes(color = continent)) +
  scale_x_log10(labels = scales::comma) +
  scale_size(labels = scales::comma)+
  ggtitle("World in 2007") +
  labs(x="GDP per capita",
       y= "life expectancy") +
  theme(text = element_text(size = 14))
 


richest_2007 <- gapminder %>%
  filter(year == 2007) %>%
  arrange(desc(gdpPercap)) %>%
  slice_head(n = 5)

view(richest_2007)
```

...

The black bubbles are a bit hard to read, the comparison would be easier with a bit more visual differentiation.

### Questions for the static figures:

1. *Answer: why does it make sense to have a log10 scale (`scale_x_log10()`) on the x axis?* (hint: try to comment it out and observe the result)
  Answer: Better visibility 

2. *Answer: In Figure 1: Who is the outlier (the richest country in 1952) far right on the x axis?* 
  Answer: Kuwait

3. *Fix Figures 1 and 2: Differentiate the **continents** by color, and fix the axis labels and units to be more legible (**Hint**: the 2.50e+08 is so called "scientific notation". You want to eliminate it.)*
  Answer: done

4. *Answer: What are the five richest countries in the world in 2007?*
  Answer: Norway, Kuwait, Singapore, United States, Ireland



##  Make it move!

The comparison would be easier if we had the two graphs together, animated. We have a lovely tool in R to do this: the `gganimate` package. Beware that there may be other packages your operating system needs in order to glue interim images into an animation or video. Read the messages when installing the package.

Also, there are *two* ways of animating the gapminder ggplot.

### Option 1: Animate using transition_states() 

The first step is to create the object-to-be-animated
```{r anim1 start}
anim <- ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop)) +
  geom_point(aes(color = continent)) +
  scale_x_log10(labels = scales::comma) +
  scale_size(labels = scales::comma)+
  ggtitle("World from 1952-2007") +
  labs(x="GDP per capita",
       y= "life expectancy") +
  theme(text = element_text(size = 14))
anim
```

...

This plot collates all the points across time. The next step is to split it into years and animate it. This may take some time, depending on the processing power of your computer (and other things you are asking it to do). Beware that the animation might appear in the bottom right 'Viewer' pane, not in this rmd preview. You need to `knit` the document to get the visual inside an html file.

```{r anim1}
anim + transition_states(year, 
                      transition_length = 1,
                      state_length = 1) + 
  ggtitle("world in:{closest_state}")

```
...

Notice how the animation moves jerkily, 'jumping' from one year to the next 12 times in total. This is a bit clunky, which is why it's good we have another option. 


### Option 2 Animate using transition_time()
This option smooths the transition between different 'frames', because it interpolates and adds transitional years where there are gaps in the timeseries data.

```{r anim2, message=FALSE, warning=FALSE}
anim2 <- ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, color = continent)) +
  geom_point() +
  scale_x_log10(labels = scales::comma) +
  scale_size(labels = scales::comma) +
  labs(x="GDP per capita", y= "life expectancy") +
  ggtitle("World in {frame_time}") +
  theme(text = element_text(size = 14)) +
  transition_time(year)

animate(anim2, nframes = 100, fps = 10, renderer = gifski_renderer())




```

The much smoother movement in Option 2 will be much more noticeable if you add a title to the chart, that will page through the years corresponding to each frame.

Now, choose one of the animation options and get it to work. You may need to troubleshoot your installation of `gganimate` and other packages 

### Tasks for the animations:

5. *Can you add a title to one or both of the animations above that will change* 
 *in sync with the animation?* *(**Hint**: search labeling for `transition_states()` and `transition_time()` functions respectively)*


6. *Can you made the axes' labels and units more readable? Consider expanding the abbreviated labels as well as the scientific notation in the legend and x axis to whole numbers. Also, differentiate the countries from different continents by color*


### Final Question
7. **Is the world a better place today than it was in the year you were born?** *Answer this question using the gapminder data.  Define better either as more prosperous, more free, more healthy, or suggest another measure that you can get from gapminder. Submit a 250 word answer with an illustration to Brightspace. Include a URL in your Brightspace submission that links to the coded solutions in Github. [Hint: if you wish to have more data than is in the filtered gapminder, you can load either the `gapminder_unfiltered` dataset or download more historical data at https://www.gapminder.org/data/ ]*

The Gapminder data that shows GDP per capita/life expectency stops in 2007. Therefore i can't use this data to show if the world is a better place than when i was born. Therefore i will calculate change in democracy from 2006-2020.

If you look at avg_democracy it will show that the median score of democracy actually went down a bit worldwide. 55.02242 in 2006 till 53.62298 in 2020. The problem with this data is that every country is equal, so the democracy score in Denmark will weight as much as India, even though the population is quite different.

So if we use this graph, it tells the world is not a better place now (at least if you like democracy). The problem is that "better" can mean a lot of things, so this one dateset is not enough to answer this question.

```{r}
  # data that shows if the world is better than when i was born
  # i use a dateset that shows chance in democracy, which starts in 2006 (close to when i was born), and ends in 2020 (close to now)



library(tidyverse)  
Democracy <-read_csv("demox_eiu.csv")

head(Democracy)
view(Democracy)
names(Democracy)

library(ggplot2)
library(dplyr)
library(tidyr)

# Convert from wide to long format
Democracy_long <- Democracy %>%
  pivot_longer(
    cols = `2006`:`2020`,      # All year columns
    names_to = "year",
    values_to = "democracy_score"
  ) %>%
  mutate(year = as.numeric(year))


# theres many countries, but we cant see their population, so lets clam them together

  change <- Democracy_long %>%
  filter(year %in% c(2006, 2020)) %>%
  pivot_wider(names_from = year, values_from = democracy_score) %>%
  mutate(change = `2020` - `2006`)

summary(change$change)
  

avg_democracy <- Democracy_long %>%
  group_by(year) %>%
  summarise(mean_score = mean(democracy_score, na.rm = TRUE))

# now lets make a graph

ggplot(avg_democracy, aes(x = year, y = mean_score)) +
  geom_line(size = 1.2) +
  geom_point() +
  scale_y_continuous(limits = c(min(avg_democracy$mean_score) - 1,
                                max(avg_democracy$mean_score) + 1),
                     n.breaks = 10) +
  theme_minimal() +
  labs(
    title = "Average Democracy Score (2006–2020)",
    x = "Year",
    y = "Average Score"
  )
```

